AAAI Press Formatting Instructions \\for Authors Using \LaTeX{} --- A Guide

### 2021-09-14 publications/iaai21/dora_outlier_detection_iaai.tex

We simulated the operational setting in which the rover has observed
targets up through mission day (sol) $s$ and the goal is to rank all
subsequent targets (after sol $s$) to inform which recent targets
merit further study.  We organized the Mars target images
chronologically and partitioned them into the corresponding ``prior''
set $D_s$ (sols \num{1343} to \num{1693}, $n=992$) and ``assessment''
set $D_{s+}$ (sols \num{1694} to \num{1703}, $n=58)$.

We collaborated with an MSL science planning team member to
independently review the targets in $D_{s+}$ and identify those that were
considered novel by the mission ($n_{outlier} = 9$).  Our goal for
this application is to assess how well the selections made by each
algorithm correlate with human novelty judgments to determine which
methods would be most suitable for informing onboard decisions about
follow-up observations.

@@

\subsubsection{Targets in Mars rover images.}
% We initialized each algorithm with $D_s$ and requested a ranking of
% targets in $D_{s+}$. 
\todo{Add table showing number (or percentage?) of novel targets in
the top $N$ selections, where $N$ = known number of novel targets in
$D_{s+}$.  And/or, report average rank of novel targets if alg ranks all
$|D_{s+}|$ targets. If time/room permits, do this for different values
of $s$.}

\subsubsection{Satellite time series for ground-truth observations.}
We used the full dataset of 6,757 samples to fit the ranking algorithms and
used the independently-labeled set of 113 points in Kenya for evaluation.
Table \ref{fig:faw_results} compares the performance of each ranking 
algorithm by plotting the number of true outliers detected out of the top
$N$ selections. We reported the area under the curve (AUC), normalized
by the total number of selections, in the legend
for each algorithm to give a quantitative comparison of the curves.
Table \ref{tab:faw_results} gives the precision at $N$ where
$N=n_{outlier}=37$. The negative sampling algorithm had the best performance
in both metrics. DEMUD, PCA, and PAE tied for the lowest precision at 
$N=n_{outlier}$ while DEMUD and PCA tied for the lowest AUC.

\begin{table}
  \caption{Precision at $N=n_{outlier}$ for three applications; the
  best result for each data set is in bold.}
  \label{tab:faw_results}
  \centering
  \begin{tabular}{l|ccc}
    \hline
    Algorithm & Earth & Planetary & Astrophysics \\
    \hline
    PCA               & 0.41 & 0.44 \\
    DEMUD             & 0.41 & 0.44 \\
    \hline
    RX                & 0.43 & ---- \\
    LRX               & ---- & 0.33 \\
    \hline
    Isolation Forest  & 0.46 & \textbf{0.56} \\
    \hline
    PAE               & 0.41 & 0.22 \\
    Neg. Sampling & \textbf{0.49} & 0.33 \\
    \hline
    Random            & 0.32 & 0.11 \\
    \hline
  \end{tabular}
\end{table}

>>>>

We simulated the operational setting in which the rover has observed
targets up through mission day (sol) $s$ and the goal is to rank all
subsequent targets (after sol $s$) to inform which recent targets
merit further study.  Our rover image data covers sols \num{1343}
to \num{1703}.  We partitioned the images chronologically to assess
outlier detection in the 10 most recent sols, using ``prior'' set
$D_{1343-1693}$ ($n=992$) and ``assessment'' set $D_{1694-1703}$
($n=58)$ for evaluation.

We collaborated with an MSL science planning team member to
independently review the targets in $D_{1694-1703}$ and identify those
that were 
considered novel by the mission ($n_{outlier} = 9$).  Our goal for
this application is to assess how well the selections made by each
algorithm correlate with human novelty judgments to determine which
methods would be most suitable for informing onboard decisions about
follow-up observations.

@@

\subsubsection{Targets in Mars rover images.}
% We initialized each algorithm with $D_s$ and requested a ranking of
% targets in $D_{s+}$.
\todo{Consider merging all three data set discussions since they are
in one table? with their own figures? The point here isn't really to
compare algs or data sets but instead to show that we CAN compare
them, using DORA.}
%
Table~\ref{tab:faw_results} compares the performance of each ranking 
algorithm by reporting the precision obtained when the number of
selections is the same as the total number of outliers.  We could not
employ RX in this case since the data dimensionality ($64 \times 64 =
4096$) exceeded the data set size.  We found that
the Isolation Forest achieved the highest precision (best outlier
detection) when allowed to select only \num{9} images.
Figure~\ref{fig:planetary_results} shows the complete (cumulative)
outlier detection performance for each algorithm when ranking
all \num{58} target images in $D_{1694-1703}$. \todo{Explain MDR the first
time it appears. }

\subsubsection{Satellite time series for ground-truth observations.}
We used the full dataset of 6,757 samples to fit the ranking algorithms and
used the independently-labeled set of 113 points in Kenya for evaluation.
Table \ref{fig:faw_results} compares the performance of each ranking 
algorithm by plotting the number of true outliers detected out of the top
$N$ selections. We reported the area under the curve (AUC), normalized
by the total number of selections, in the legend
for each algorithm to give a quantitative comparison of the curves.
Table \ref{tab:faw_results} gives the precision at $N$ where
$N=n_{outlier}=37$. The negative sampling algorithm had the best performance
in both metrics. DEMUD, PCA, and PAE tied for the lowest precision at 
$N=n_{outlier}$ while DEMUD and PCA tied for the lowest
AUC. \todo{For completeness, Explain why no LRX here.}

\begin{table}
  \caption{Precision at $N=n_{outlier}$ for three applications; the
  best result for each data set is in bold.}
  \label{tab:faw_results}
  \centering
  \begin{tabular}{l|ccc}
    \hline
    Algorithm & Astrophysics & Earth & Planetary \\
    \hline
    PCA               & & 0.41 & 0.44 \\
    DEMUD             & & 0.41 & 0.44 \\
    \hline
    RX                & & 0.43 & ---- \\
    LRX               & & ---- & 0.33 \\
    \hline
    Isolation Forest  & & 0.46 & \textbf{0.56} \\
    \hline
    PAE               & & 0.41 & 0.22 \\
    Neg. Sampling & & \textbf{0.49} & 0.33 \\
    \hline
    Random            & & 0.32 & 0.11 \\
    \hline
  \end{tabular}
\end{table}
